---
title: "data_cleaning"
format: html
editor_options: 
  chunk_output_type: console
---
This Qmd contains all cleaning of the raw data before ingesting into a relational database.

Load libraries
```{r}
library(here)
library(tidyverse)
library(lubridate)
```

read in data
```{r}
org_keylist <- read_csv(here("data", "raw", "Benthic organism keylist 2024.csv"))
cpue <- read_csv(here("data", "raw", "DWR Benthic CPUE data 1975-2024.csv"))
site <- read_csv(here("data", "raw", "DWR benthic monitoring, active and historic site locations.csv"))
```

Remove duplicated columns from CPUE, since these columns will be represented in other appropriate tables.
```{r}
cpue_clean <- cpue |> 
  select(-c(Class_level, Order_level, Family_level, Phylum, Genus, Species, Common_name, Location, Latitude, Longitude))
```

Adjust column types in CPUE data
```{r}
cpue_clean <- cpue_clean |> 
  mutate(Date = as_date(Date),
         OrganismCode = as.character(OrganismCode))
```

Check for NAs to define in schema if necessary.
```{r}
# CPUE table
sum(is.na(cpue_clean$Date))
sum(is.na(cpue_clean$Station))
sum(is.na(cpue_clean$Year))
sum(is.na(cpue_clean$Month))
sum(is.na(cpue_clean$OrganismCode))
sum(is.na(cpue_clean$MeanCPUE))
sum(is.na(cpue_clean$TotalGrabs))

# Organism keylist table
sum(is.na(org_keylist$OrganismCode))
sum(is.na(org_keylist$Phylum))
sum(is.na(org_keylist$Class_level))
sum(is.na(org_keylist$Order_level)) # 8 NAs
sum(is.na(org_keylist$Family_level))
sum(is.na(org_keylist$Genus))
sum(is.na(org_keylist$Species))
sum(is.na(org_keylist$Common_name)) # 8 NAs also

# Site table
sum(is.na(site$StationCode))
sum(is.na(site$Location))
sum(is.na(site$Latitude))
sum(is.na(site$Longitude))
sum(is.na(site$Period_of_Record_From))
sum(is.na(site$Period_of_Record_To))
sum(is.na(site$Missing)) # 43 NAs
sum(is.na(site$Status))

```

Remove NAs from organism keylist, as we do not want rows where organisms were defined to the species level but missing order information.
```{r}
org_keylist_clean <- org_keylist |> 
  filter(!is.na(Order_level)) |> 
  mutate(OrganismCode = as.character(OrganismCode))
```

Adjust column types in site table
```{r}
site_clean <- site |> 
  mutate(Period_of_Record_From = as.character(Period_of_Record_From))
```

Check for non-unique primary key in site table, as there are issues ingesting.
```{r}
site_clean |> 
  count(StationCode) |> 
  filter(n > 1)
```

There are 2 values of D4-L with slightly different lat and long, likely the station location changed at some point during sampling. For simplification, I will just remove one row as I do not see any information in metadata.
```{r}
site_clean <- site_clean[-5, ]
```

Save processed data
```{r}
write_csv(cpue_clean, file.path(here("data", "processed", "cpue.csv")))
write_csv(org_keylist_clean, file.path(here("data", "processed", "org_keylist.csv")))
write_csv(site_clean, file.path(here("data", "processed", "site.csv")))
```

